---
title: "Mini Literate Analysis"
format: html
---

## Introduction

### About this data

This data comes from the North Pole Environmental Observatory Bottle Chemistry Project, which includes measurements about the seawater chemistry, taken from 2000-2014. It was downloaded from the [Artic Data Center](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X) on September 13, 2023.

### Setup - call all libraries at the top of your code

```{r}
library(readr) #part of tidyverse suite of packages
```

### Read in data

```{r}
#| warning: false 
#don't add space between {r} and #| or else code won't run 

getwd() #tells you where you're starting point is

bg_chem <- read_csv ("data/BGchem2008data.csv")
#get warnings when getting data but don't want them to appear in report so use #| warning:false at top of the code

```

-   Exploring my data

```{r}
#| echo: false
#| eval: false
#information of dataframe
## Prints the column names of my data frame
colnames(bg_chem)

## General structure of the data frame - shows class of each column
str(bg_chem)

## First 6 lines of the data frame
head(bg_chem)

## Summary of each column of data
summary(bg_chem)

## Prints unique values in a column (in this case Date)
unique(bg_chem$Date)

```


## Analysis

### Calculate summary statistics
```{r}
nitrate <- mean (bg_chem$NO3)

nitrite <- mean (bg_chem$NO2)

amm <- mean (bg_chem$NH4)

phos <- mean (bg_chem$P)
```


### Calculate the Redfield Ratio

```{r}
ratio <- (nitrate + nitrite + amm)/ phos
ration1 <- sum(nitrate + nitrite + amm)/phos
```


### Plot Redfield Ratio
```{r}
#| echo: false
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)

# p = x-axis
# nitrogen components = y-axis
```


## Conclusion

The Redfield Ratio for this dataset is approximately `r round(ratio)`.
